---
title: "CSC 235 Final Project: Data Cleanup"
author: "Allie"
date: "5/18/2021"
output: 
  html_document:
    theme: united
    highlight: tango
    toc: yes
    toc_float: yes
---

# Read in Data

```{r global_options, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
```

```{r packages}
library(dplyr)      # run install.packages("package_name")
library(stringr)    # in console to install missing packages
library(stringi)
```


## Read in March data

```{r}
mar19 <- read.csv("original_data/original-2019-03-march.csv", 
                  na.strings=c(""," ",NA))

mar20 <- read.csv("original_data/original-2020-03-march.csv", 
                  na.strings=c(""," ",NA))
```


## Read in April data

```{r}
apr19 <- read.csv("original_data/original-2019-04-april.csv", 
                  na.strings=c(""," ",NA))

apr20 <- read.csv("original_data/original-2020-04-april.csv", 
                  na.strings=c(""," ",NA))
```


# Delete Unnecessary Columns and Filter by 3 Major New York Airports

```{r}
# dataframes to include
data_frames <- c("mar19", "mar20", "apr19", "apr20")
# airports to filter by
nyAirports <- c("EWR", "JFK", "LGA")

# columns to delete
colsDel <- c("X", "DUP", "ORIGIN_CITY_NAME", "ORIGIN_STATE_ABR",
             "ORIGIN_STATE_NM", "DEST_CITY_NAME", "DEST_STATE_ABR",
             "DEST_STATE_NM", "DEP_DELAY_NEW", "TAXI_OUT", "WHEELS_OFF",
             "WHEELS_ON", "TAXI_IN", "ARR_DELAY_NEW", "CRS_ELAPSED_TIME",
             "ACTUAL_ELAPSED_TIME", "AIR_TIME", "DISTANCE", "FIRST_DEP_TIME",
             "TOTAL_ADD_GTIME", "DIV_ACTUAL_ELAPSED_TIME", "BRANDED_CODE_SHARE",
             "TAIL_NUM", "OP_CARRIER_FL_NUM", "CANCELLATION_CODE", 
             "DIV_AIRPORT_LANDINGS", "DIV_DISTANCE", "DEP_TIME", "ARR_TIME")
```

```{r}
colsStr <- paste0("-", str_c(colsDel, collapse=", -"))

for(fr in data_frames) {
  eval(parse(text=paste0(fr," <- select(",fr, ", c(",colsStr,"))")))
  eval(parse(text=paste0(fr," <- ",fr,
                         " %>% filter(DEST %in% nyAirports | ORIGIN %in% nyAirports)")))
}
```


# Combine dataframes

```{r}
flNY <- rbind(mar19, mar20, apr19, apr20)
```


# Format Existing Columns

## Rename columns

```{r}
names(flNY) <- str_to_lower(names(flNY))
```

```{r}
flNY <- flNY %>%
  rename(weekday = day_of_week,
         date = fl_date,
         network_airline = mkt_unique_carrier,
         operating_airline = op_unique_carrier)
```

## Format time columns

```{r}
time_cols <- c("crs_dep_time", "crs_arr_time")

formatTime <- function(time_col) {
  x <- as.character(time_col)
  x <- str_pad(x, 4, side="lef", pad="0")
  stri_sub(x, 3, 2) <- ":"
  return(x)
}

for(col in time_cols) {
  eval(parse(text=paste0("flNY$",col," <- formatTime(flNY$",col,")")))
}
```

## Format `cancelled`, `diverted`, and `div_reached_dest` columns

```{r}
flNY$cancelled <- as.logical(flNY$cancelled)
flNY$diverted <- as.logical(flNY$diverted)
flNY$div_reached_dest <- as.logical(flNY$div_reached_dest)
```


# Create New Columns

## `month.name` column

```{r}
lookup_days <- read.csv("original_data/lookup_tables/L_MONTHS.csv_")
getMonth <- lookup_days$Description
names(getMonth) <- lookup_days$Code

flNY$month.name <- getMonth[flNY$month]
```


## `weekday.name` columns

```{r}
lookup_days <- read.csv("original_data/lookup_tables/L_WEEKDAYS.csv_")
getWeekday <- lookup_days$Description
names(getWeekday) <- lookup_days$Code

flNY$weekday.name <- getWeekday[flNY$weekday]
```


## `origin.name` and `dest.name` columns

```{r}
lookup_airport <- read.csv("original_data/lookup_tables/L_AIRPORT.csv_")
getAirport <- str_split_fixed(lookup_airport$Description, ": ", 2)[,2]
names(getAirport) <- lookup_airport$Code

flNY$origin.name <- getAirport[flNY$origin]
flNY$dest.name <- getAirport[flNY$dest]
```

## `network_airline.name` and `operating_airline.name` columns

```{r}
lookup_carrier <- read.csv("original_data/lookup_tables/L_UNIQUE_CARRIERS.csv_")
getAirline <- lookup_carrier$Description
names(getAirline) <- lookup_carrier$Code

flNY$network_airline.name <- getAirline[flNY$network_airline]
flNY$operating_airline.name <- getAirline[flNY$operating_airline]
```


## `dep_delay.c` and `arr_delay.c` columns

```{r}
Fdep_delay.c <- function(row) {
  if(flNY$cancelled[row] == TRUE) {
    return(NA)
  } else { # not cancelled
    return(flNY$dep_delay[row])
  }
}

Farr_delay.c <- function(row) {
  if(flNY$cancelled[row] == TRUE) {
    return(NA)
  } else if(flNY$diverted[row] == TRUE) {
    if(flNY$div_reached_dest[row] == FALSE) {
      return(NA)
    } else { # flNY$div_reached_dest[row] == TRUE
      return(flNY$div_arr_delay[row])
    }
  } else { # not diverted or cancelled
    return(flNY$arr_delay[row])
  }
}

flNY$dep_delay.c <- sapply(1:nrow(flNY), Fdep_delay.c)
flNY$arr_delay.c <- sapply(1:nrow(flNY), Farr_delay.c)
```


## `dep_delay.info` and `arr_delay.info` columns

```{r}
delayCat <- c("Early", "On Time", "Delay 0:15 to 0:29", "Delay 0:30 to 0:44",
              "Delay 0:45 to 0:59", "Delay 1:00 to 1:14", "Delay 1:14 to 1:29", 
              "Delay 1:30 to 1:44", "Delay 1:45 to 1:59", "Delay 2:00 to 2:14",
              "Delay 2:15 to 2:29", "Delay 2:30 to 2:44", "Delay 2:45 to 2:59",
              "Delay 3:00 and above", "Cancelled or Diverted")

Fdelay.info <- function(delay) {
  if(is.na(delay)) {
    return(delayCat[15])
  } else if(delay < 0) {
    return(delayCat[1])
  } else if(delay < 15) {
    return(delayCat[2])
  } else if(delay < 30) {
    return(delayCat[3])
  } else if(delay < 45) {
    return(delayCat[4])
  } else if(delay < 60) {
    return(delayCat[5])
  } else if(delay < 75) {
    return(delayCat[6])
  } else if(delay < 90) {
    return(delayCat[7])
  } else if(delay < 105) {
    return(delayCat[8])
  } else if(delay < 120) {
    return(delayCat[9])
  } else if(delay < 135) {
    return(delayCat[10])
  } else if(delay < 150) {
    return(delayCat[11])
  } else if(delay < 165) {
    return(delayCat[12])
  } else if(delay < 180) {
    return(delayCat[13])
  } else { #delay >= 180
    return(delayCat[14])
  }
}

flNY$dep_delay.info <- sapply(flNY$dep_delay.c, Fdelay.info)
flNY$arr_delay.info <- sapply(flNY$arr_delay.c, Fdelay.info)
```


# Write final dataframe to .csv file

```{r}
write.csv(flNY, "flight_delay_shiny_app/files/NY-flights.csv", row.names=FALSE)
```




